Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-02 manuscript/manuscript.tex
updated ms discussion

%We visualize these covariances in Figure \label{figure-1} (A), which
%depicts the temporal covariances through time, for each of the five
%off-diagonal rows temporal covariance matrix. Each row represents the temporal
%covariance $\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between some initial
%reference generation $s$ (the row of the matrix), and some later timepoint $t$
%(the column of the matrix). For each row, the covariances at first are
%positive, and then decay towards zero as expected when directional selection
%affects linked variants' frequency trajectories until ultimately linkage
%disequilibrium and additive genetic variance for decay
%\parencite{Buffalo2019-io}. Note that per replicate, the signal is a bit
%noisier; see Supplementary Figures XXX.

%Since the \textcite{Barghi2019-qy} study sequenced each
% replicate population every ten generations, we calculate the temporal
% covariance matrix over ten-generation allele frequency changes, which we write
% as $\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$. The diagonal elements of
% this temporal covariance matrix are the sum of all variances in allele
% frequency changes and pairwise covariances, e.g. element $\mathbf{Q}_{t,t} =
% \sum_{i=t}^{9} \var(\Delta p_i) + 2\sum_{i \ne j} \cov(\Delta p_i, \Delta
% p_j)$. Each off-diagonal element $\mathbf{Q}_{t,s}, \; t \ne s$ contains the
% sum of pairwise covariances, $\mathbf{Q}_{t,s} = \sum_{i=t}^{t+9}
% \sum_{j=s}^{s+9} \cov(\Delta p_i, \Delta p_j)$ (see Supplementary Material
% Section \ref{supp:barghi-covs} for more details). These off-diagonal
% covariances are expected to be zero when only drift is acting in the
% population, as only linked selection can create covariance between allele
% frequency changes \parencite{Buffalo2019-io}. Furthermore, since we cannot
% directly observe the temporal covariances between adjacent timepoints, but only
% the sums of temporal covariances, we expect the off-diagonal elements of
% $\mathbf{Q}$ to be weak. This is because the magnitude of a temporal covariance
% $\cov(\Delta p_i, \Delta p_j)$ is governed by the level of additive genetic
% variance and linkage disequilibrium that persists after $|i-j|$ generations of
% decay, which in \textcite{Barghi2019-qy}'s study design is on average 10
% generations.

@@

%\gc{TODO GIVE conversion \% var due to linked selection. }
Second, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Castro2019-uk,Marchini2014-de}. This study includes two independent selection lines,
Longshanks 1 and 2 (LS1 and LS2), and an unselected control line (Ctrl).
Consequently, this selection experiment offers a useful control to test our
between-replicate covariances: we expect to see positive between-replicate
covariance in the comparison between the two Longshanks selection lines, but
not between the two pairwise comparisons between the control line and the two
Longshanks lines. We find that this is the case (gray confidence intervals in
Figure \ref{fig:figure-2} C), with the two Longshanks comparisons to the
control line not being significantly different from zero, while the comparison
between the two Longshanks line is statistically significantly different from
zero (CIs [0.013, 0.037]).

@@

%We find that for covariances 2 timepoints apart ($k=2
%\times 10$ generations), the $20\%$ right tail is inflated by a of 1.3 (95\% CI
%[1.29, 1.32]), consistent with linked selection creating positive covariance in
%allele frequency change. Under constant directional selection, we expect these
%positive covariances between increasingly distant timepoints to decay to zero
%as additive genetic variance is exhausted by selection and linkage
%disequilibrium decays. Instead we find that for covariances 4 timepoints apart
%($k=4 \times 10$ generations), the left tail of negative covariances is
%inflated by a factor of 1.19 (95\% CI [1.17, 1.21]).

@@

Our analyses reveal that a sizable proportion of allele frequency
change in these populations is due to the (indirect) action of
selection. Capitalizing on replicated designs, we characterized the extent to which
convergent selection pressures lead to parallel changes in allele frequencies
across replicate populations, and found that a reasonably proportion
of the response is shared across short timescales. These likely
represent substantial under-estimates of the contribution of linked
selection because ....

These estimates of the contribution of selection could be refined by using
patterns of LD and recombination which would allow us to more fully
parameterize a linked-selection model of temporal allele frequency change
\parencite{Buffalo2019-io}. The basic prediction is that regions of higher
linkage disequilibrium and lower recombination should have greater temporal
autocovariance than regions with lower LD and higher recombination. However,
one limitation of these pooled sequence datasets is that none of the studies we
reanalyzed estimated linkage disequilibria data for the evolved populations.
There are LD data for a natural population of \emph{D. simulans}
\parencite{Signor2018-wg},  we did not find a relationship between temporal
covariance and LD.  We believe this is driven by the idiosyncratic nature of LD
in evolve-and-resequence populations, which often extends over large genomic
distances \parencite{Nuzhdin2013-gf,Kelly2019-dc}. Future studies complete with
LD data and recombination maps would allow one to disentangle the influence of
closely linked sites from more distant sites in causing temporal
autocovariance, and allow the fitting of more parameteric model to estimate
population parameters such as the additive genetic variation for fitness
directly from temporal gneomic data alone  \parencite{Buffalo2019-io}.

Our primary focus here has been on evolution in laboratory
populations. It is unclear whether we should expect a similar impact
of selection in natural populations. In some of these experiments, selection pressures may
have been stronger or more sustained that in natural populations REF.
Conversely, they these lab populations were maintained at very small
population sizes, small effective population sizes 300, 450, and 45 for the
\textcite{Kelly2019-dc}, and \textcite{Castro2019-uk} studies
respectively, which will amplify the role of genetic drift. The
advantage of lab experiments is that they are closed populations, in
natural populations temporal covariance could also arise from
the systematic migration of alleles from differentiated
populations. To these methods to natural populations will require
either populations that are reasonably closed to migration, or for the effect of migration to
be accounted for possibly either by knowledge of allele frequencies in source
populations or the identification of migrant individuals.

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to estimate the quantify
the impact of linked selection in have suggested that XXX to XXX of allele
frequency change over long time periods is driven by selection
\parencite{Mcvicker,Comeron,Elyashiv2016-vt,Russ,Coop}. Thus there is reason to
be hopeful that we could detect this from temporal data, which would help to
resolve the time-scales that linked selection act over. In our reanalysis of
the \textcite{Barghi2019-qy} study, we find evidence of complex linked
selection dynamics, with selection pressures flipping over time due to either
environmental change, the breakup of epistatic combinations or advantageous
haplotypes. Such patterns would be completely obscured in samples only from
contemporary populations, thus we can hope to have a much richer picture of
linked selection as temporal sequencing becomes more common.

@@

% \begin{align}
%   \mathbf{\Delta F} &= 
%                     &\begin{bmatrix} 
%     p_{1, 0, 0} - p_{0, 0, 0} & p_{2, 0, 0} - p_{1, 0, 0} & \ldots & p_{1, 1, 0} - p_{0, 1, 0} & p_{2, 1, 0} - p_{1, 1, 0} & \ldots & p_{T+1, R, 0} - p_{T, R, 0}  \\
%     p_{1, 0, 1} - p_{0, 0, 1} & p_{2, 0, 1} - p_{1, 0, 1} & \ldots & p_{1, 1, 1} - p_{0, 1, 1} & p_{2, 1, 1} - p_{1, 1, 1} & \ldots & p_{T+1, R, 1} - p_{T, R, 1}  \\
%     \vdots & \vdots & \ddots & \vdots & \vdots & \ddots & \vdots  \\
%     p_{1, 0, L} - p_{0, 0, L} & p_{2, 0, L} - p_{1, 0, L} & \ldots & p_{1, 1, L} - p_{0, 1, L} & p_{2, 1, L} - p_{1, 1, L} & \ldots & p_{T+1, R, L} - p_{T, R, L}  \\
%   \end{bmatrix} 
% \end{align}

@@

%Additionally, in some study designs, a single timepoint is shared for the
%initial generation across replicates. In this case, the sampling noise is
%shared between 
% TODO

@@

@@

% ======= We were concerned that this might be due to a flaw in our methods. To
% address this concern, we first ensured that our temporal-covariance matrix
% bias correction was working as expected. We find that it corrects the
% relationship between depth and both variance and covariance (Supplementary
% Figure \ref{suppfig:bergland-correction}) as expected. We then re-analyzed
% the data of \parencite{Bergland2014-ij} using the same seasonal fluctuating
% model as the original paper. This model is a Binomial logit-linked GLM fit
% per-locus, where the Spring/Fall seasons are encoded as a dummy variable, and
% are regressed on the frequency data. We use the same binomial weighting
% procedure as \textcite{Bergland2014-ij}, where the weights are determined by
% the effective number of chromosomes, $N_{eff} = (2 n_t d_t - 1) / (2 n_t +
% d_t)$ ($n_t$ and $d_t$ are the number of diploid individuals and the read
% depth at timepoint $t$, respectively). We fit this model on all loci marked
% as used in the VCF provided with the \textcite{Bergland2014-ij} study
% (doi:10.5061/dryad.v883p).  Overall, our p-values for the Wald test for each
% locus closely match those of the original paper (Pearson correlation
% coefficient 0.98, p-value < $2.2 \times 10^{-16}$; see Supplementary Figure
% \ref{suppfig:bergland-pvalue-comparison} A), and the histograms of the
% p-values are nearly identical (Supplementary Figure
% \ref{suppfig:bergland-pvalue-comparison} B).  >>>>>>>
% d95c3549a10dc0037f40a54cb8b41f9d412ed1ed

@@

%\gc{ADD MORE HERE} TODO

%\gc{Note that this analysis does not rule out the idea that prior
 % analyses detected {\emph
  %  some} loci showing a pattern of fluctuating selection. Indeed
  %\textcite{Bergland2014-ij} provided a number of analyses that
  %suggested that their loci might be enriched for {\bf XXX}. However,
  %this reanalysis suggests that we currently lack genome-wide evidence
%of many loci experiencing fluctuating selection in {\it in Drosophila
 % melanogaster}. }

>>>>

@@

Second, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Castro2019-uk,Marchini2014-de}. This study includes two independent selection lines,
Longshanks 1 and 2 (LS1 and LS2), and an unselected control line (Ctrl).
Consequently, this selection experiment offers a useful control to test our
between-replicate covariances: we expect to see positive between-replicate
covariance in the comparison between the two Longshanks selection lines, but
not between the two pairwise comparisons between the control line and the two
Longshanks lines. We find that this is the case (gray confidence intervals in
Figure \ref{fig:figure-2} C), with the two Longshanks comparisons to the
control line not being significantly different from zero, while the comparison
between the two Longshanks line is statistically significantly different from
zero (CIs [0.013, 0.037]).

@@

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because \vb{the studies we have reanalyzed
do not sequence the population each generation, preventing us from including
the effects of stronger correlations between adjacent generations.
Furthermore, estimation methods are intentionally conservative, further
leading our estimates to be closer to a lower bound of the effects of linked
selection than a measure of central tendency.}

These estimates of the contribution of selection could be refined by using
patterns of LD and recombination which would allow us to more fully
parameterize a linked-selection model of temporal allele frequency change
\parencite{Buffalo2019-io}. The basic prediction is that regions of higher
linkage disequilibrium and lower recombination should have greater temporal
autocovariance than regions with lower LD and higher recombination. However,
one limitation of these pooled sequence datasets is that none of the studies we
reanalyzed estimated linkage disequilibria data for the evolved populations.
\vb{While} there are LD data for a natural population of \emph{D. simulans}
\parencite{Signor2018-wg},  we did not find a relationship between temporal
covariance and LD.  We believe this is driven by the idiosyncratic nature of LD
in evolve-and-resequence populations, which often extends over large genomic
distances \parencite{Nuzhdin2013-gf,Kelly2019-dc}. Future studies complete with
LD data and recombination maps would allow one to disentangle the influence of
closely linked sites from more distant sites in causing temporal
autocovariance, and allow the fitting of more parametric models to estimate
population parameters such as the additive genetic variation for fitness
directly from temporal genomic data alone \parencite{Buffalo2019-io}.

Our primary focus here has been on evolution in laboratory populations. It is
unclear whether we should expect a similar impact of selection in natural
populations. In some of these experiments, selection pressures may have been
stronger or more sustained that in natural populations
\vb{\parencite{Hendry1999-zu,Hairston2005-ga}}. Conversely, these lab
populations were maintained at very small population sizes, small effective
population sizes 300, 450, and 45 for the \textcite{Kelly2019-dc}, and
\textcite{Castro2019-uk} studies respectively, which will amplify the role of
genetic drift. The advantage of lab experiments is that they are closed
populations, in natural populations temporal covariance could also arise from
the systematic migration of alleles from differentiated populations.
\vb{Adapting} these methods to natural populations will require either
populations that are reasonably closed to migration, or for the effect of
migration to be accounted for possibly either by knowledge of allele
frequencies in source populations or the identification of migrant individuals.

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to estimate the quantify
the impact of linked selection in have suggested a 19\% to 89\% \vb{reduction
in genome-wide diversity} over long time periods is driven by \vb{linked
selection}
\parencite{McVicker2009-ax,Elyashiv2016-vt,Corbett-Detig2015-gt,Coop2016-gx,Comeron2014-nh}
Thus there is reason to be hopeful that we could detect this from temporal
data, which would help to resolve the timescales that linked selection act
over. In our reanalysis of the \textcite{Barghi2019-qy} study, we find evidence
of complex linked selection dynamics, with selection pressures flipping over
time due to either environmental change, the breakup of epistatic combinations
or advantageous haplotypes. Such patterns would be completely obscured in
samples only from contemporary populations, thus we can hope to have a much
richer picture of linked selection as temporal sequencing becomes more common,
\vb{perhaps allowing us to observe the effects of ecological dynamics in
genomic data \parencite{Hairston2005-ga}.}

@@

@@

@@

\subsection{Proportion of Variance Explained by Linked Selection across Replicates}

We define a statistic similar to $G$ for estimating the proportion of allele
frequency change common between two replicate populations due to linked
selection. Covariance in allele frequency change between two replicate
populations is due to convergent selection pressure selecting haplotypes shared
between the two replicate populations, which acts to perturb linked neutral
variation in parallel way. Similar to \textcite{Buffalo2019-io} equation (1).
We can similarly partition the allele frequency change,

\begin{align}
  \label{eq:}
  \Delta p_t &= \underbrace{\Delta_{_N} p_t  + \Delta_{_M} p_t}_\text{drift} + \underbrace{\Delta_{_S} p_t}_\text{shared} + \underbrace{\Delta_{_U} p_t}_\text{unique}
\end{align}

where $\Delta_{_N} p_t$ and $\Delta_{_M} p_t$ are the allele frequency changes
due to non-heritable variation in fitness (e.g. environmental differences in
family size) and Mendelian segregation. In \textcite{Buffalo2019-io} allele
frequency changes due to heritable fitness variation are captured in the term
$\Delta_{_H} p_t$; here we further partition this into heritable frequency
changes shared between replicates ($\Delta_{_S} p_t$) and those unique to a
replicate ($\Delta_{_H} p_t$). By definition,

\begin{align}
  \cov(\Delta p_{s,A}, \Delta p_{t,B}) &= \cov(\Delta_{_S} p_{s,A}, \Delta_{_S} p_{t,B}), \;\; A \ne B\\
  \cov(\Delta p_{s,A}, \Delta p_{t,A}) &= \cov(\Delta_{_S} p_{s,A} + \Delta_{_U} p_{s,A}, \Delta_{_S} p_{t,A} + \Delta_{_U} p_{t,A}) 
\end{align}

Then, we define a replicate covariance analog of $G$, $G_R$,

% \begin{align}

% \end{align}

@@

@@
